Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2023-01-16 manuscripts/main.tex
last of dan's edits
latest drafts

% load packages
\usepackage[utf8]{inputenc}
\usepackage[a4paper, margin=0.8in]{geometry}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{xr-hyper} % for cross-referencing between documents
\usepackage{hyperref}
\usepackage{float}
\usepackage{amsmath}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[flushmargin]{footmisc}
\usepackage[flushleft]{threeparttable}
\usepackage{array}
\usepackage[font=small, labelfont=it]{caption}
\usepackage{booktabs} % for pretty tables

@@

% specific images path relative to the main .tex file 
\graphicspath{ {./figures/} }

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1]{Samuel Zorowitz}
\author[1,2]{Yael Niv}
\author[1,3]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, USA}
\affil[2]{Department of Psychology, Princeton University, USA}
\affil[3]{School of Psychological Sciences, Monash University, Australia}

@@

In recent years, online labour markets (e.g. Amazon Mechanical Turk, Prolific, Cloudresearch) have become increasingly popular as a source of research participants in the behavioral sciences \cite{stewart2017crowdsourcing}, in no small part due to the ease with which these services allow for recruitment of large, diverse samples. The advantages of online data collection have also begun to be recognized in psychiatric research \cite{chandler2016conducting}, where this method offers several distinct advantages over traditional approaches to participant recruitment. The ability to assess psychiatric symptom severity in large general-population samples makes possible large-scale transdiagnostic analysis \cite{gillan2016taking, rutledge2019machine}, and facilitates recruitment from difficult-to-reach participant populations \cite{strickland2019use}. Online labour markets also facilitate re-recruitment, making them an attractive option for validating the psychometric properties of assessment tools \cite{enkavi2019large} or studying clinical processes longitudinally \cite{kothe2019retention}.

@@

Here we wish to draw special attention to an underappreciated feature of psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey attentively \cite{chandler2020participant, ophir2020turker, king2018random}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. In two new datasets from two separate online labour marketplaces, we found that screening based on task behavior fails to completely identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept datasets that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms that are abolished when data are screened more thoroughly.

@@

We hypothesise that spurious behavior-symptom correlations may emerge due to a mean-shift in the average level of symptom endorsement in participants engaging in C/IE responding relative to attentive participants. In turn, a mean-shift is expected to occur when the overall rate of symptom endorsement is low; that is, comparably higher scores are more likely for C/IE participants responding at random on a right skewed questionnaire. In line with our predictions, the average level of symptom endorsement was noticeably exaggerated in C/IE-responding participants for the symptom measures where symptom scores were most positively-skewed (7-up, 7-down, GAD-7; see Figure \ref{fig:distributions}). In contrast, where there was higher rates of symptom endorsement overall, the distributions of symptom scores between the two groups of participants were less noticeably distinct. Permutation testing confirmed that observed mean-shifts in symptom scores for C/IE participants were statistically significant for the majority of symptom measures (Table \ref{tab:validation}).

@@

Second, we quantified the degree to which participants responded to self-report symptom surveys in a stereotyped fashion; that is, we determined if participants exhibited patterns in their responses that were independent of the contents of the survey items. We fit a random-intercept item factor-analysis model \cite{maydeu2006random} to self-report data (see Methods), and for each participant we estimated an intercept parameter that quantified their bias towards using responses on the left or right side of the response scale, regardless of what that response signifies for a particular self-report measure (e.g., low on one symptom scale versus high on another). We observed a credible difference between the average value of this intercept for the two groups ($\Delta \text{intercept} = 0.67, 95\% \ \text{HDI} = [-0.78, -0.55]$), such that C/IE participants were biased towards using the right-half of survey response options. This translates to a tendency to endorse \emph{more severe} symptoms on the 7-up/7-down and GAD-7 scales (where the rightmost options indicate greater frequency of symptoms) but \emph{less extreme} symptoms or personality traits on the SHAPS and BIS (where the rightmost options indicate lower frequency of symptoms or personality traits) despite these inventories measuring strongly correlated constructs (i.e. depression and anhedonia, anxiety and behavioral inhibition).

@@

\begin{figure}[!t]
\includegraphics[width=17cm]{../figures/main_03a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. Only statistically significant correlations are shown ($p<0.05$ not corrected for multiple comparisons; all signed correlations are shown in the Supplement). Black Xs indicate significant correlations ablated under screening. No Screening = no exclusions; Accuracy Only = exclusions based on chance-level performance in the reversal-learning task; Infrequency Only = exclusions based on invalid or improbable responses to infrequency items; Both Types = exclusions based on the previous two measures. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
\label{fig:correlations}
\end{figure}

@@

Next, we investigated how spurious correlations depended on sample size. To do so, we performed a bootstrapping analysis where we held fixed the proportion of participants engaging in C/IE responding (i.e. 5\%, 10\%, 15\%, 20\%) and increased the total number of participants. Across all analyses, we measured the correlation and between the 7-down depression scale and learning-rate asymmetry ($\kappa$), which we previously identified as likely exhibiting a spurious association. (The following results are not specific to learning-rate asymmetry and generalize to other pairs of variables; Figure \ref{fig:bootstrap_s1}).

The outputs of the bootstrapping analysis are presented in Figure \ref{fig:bootstrap}. We found that, although estimated correlation magnitudes were independent of sample size (x-axis, left panel), the absolute magnitude of the behavior-symptom correlation increased with the proportion of C/IE participants (different coloured circles, left panel). Crucially, we found false-positive rates for spurious correlations \emph{increased} with increases in sample size in our data for all but the smallest rates of C/IE responding (right panel). This runs counter to a common assumption that larger sample sizes are protective against spurious correlations because they serve to mitigate measurement error. Although this assumption is correct for unsystematic measurement error, it no longer holds in the regime of systematic measurement error (where larger sample sizes reduce the variance of estimates, but do not alter their bias). Instead, our results suggest that, except for low rates of C/IE responding, the false-positive rate for behavioral-symptom correlations will become increasingly inflated as the sample size increases.

@@

\subsection{Pattern of results generalizes to alternative tasks, self-report measures, and quality assurance protocols}

One possible concern with the results presented so far is that they are specific to one specific instantiation of our experimental design. For example, with more stringent quality assurance protocols during participant recruitment, or perhaps a different task or set of self-report measures, one might wonder if spurious correlations would remain such a threat.

To evaluate the generalizability of our findings, we therefore conducted a conceptual replication experiment in which an independent sample of N=393 participants (N=193 from MTurk, N=200 from Prolific) completed a more difficult cognitive task, the two-step task \cite{daw2011model}, and an alternate set of self-report measures (see Appendix B for complete details). Importantly, participants were recruited \textit{after} CloudResearch and Prolific implemented new protocols to improve data quality on their respective platforms. Moreover, participants completed self-report symptom measures as before, but also personality measures with no hypothesized relationship to model-based planning behavior on the two-step task.

For the sake of brevity, only the main pattern of finding are reported here (the complete results are reported in Appendix B). In the replication sample, 55 out of 393 participants (14\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report measures. This is roughly two-thirds of the number of participants who were flagged for C/IE responding in the original study, suggesting that the newer quality assurance protocols are at least partially effective. In the self-report symptom measures, we replicated the finding that total scores were noticeably exaggerated in participants suspected of C/IE responding only for the symptom measures where overall rates of symptom endorsement were the lowest. Similarly, we again found that task and self-report screening measures showed low correspondence; that is, excluding participants on the basis of poor behavioral performance was ineffective for identifying and removing participants who engaged in C/IE responding on self-report surveys. Finally, with no exclusions applied, we observed spurious correlations between performance on the two-step task and total scores, only for the self-report measures (both symptom and personality measures) with a mean-shift in scores between attentive participants and participants suspected of C/IE responding. In contrast with our original findings, however, we found that excluding participants based on self-report \textit{or} task screening measures was sufficient to ablate these spurious correlations.

In sum, we replicated most of the main findings from the original study in an independent sample of participants completing a novel set of tasks and self-report measures. Although we found that screening on task behavior was sufficient to protect against spurious correlations in the replication sample, it is difficult to generalize and predict when or why this might be the case for other datasets. As such, screening for C/IE responding in both task and self-report measures still remains the best approach to protect oneself against the possibility of spurious correlations.

\subsection{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}

One major concern with performing rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias \cite{elwert2014endogenous}. That is, to this point we have treated the tendency towards C/IE responding as independent from psychopathology. However, to the extent that C/IE responding reflects lack of motivation \cite{barch2015mechanisms}, avoidance of effort \cite{cohen2001impairments, culbreth2016negative}, or more frequent lapses of attention \cite{kane2016individual, robison2017neurotic}, one might hypothesise a true underlying association between psychopathology and careless responding in online studies. In such cases, it is plausible that rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants.

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Appendix C for complete details). Specifically, a total of N=45 psychiatric patients and N=20 healthy controls completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. (The psychiatric patients were confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview.) In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) psychiatric participants were flagged for C/IE responding.

Using these data, we computed pairwise Bayes factors comparing three candidate models: attention check failure rates are equal between healthy and psychiatric participants (M1); failure rates are greater in psychiatric patients (M2); and failure rates are greater in healthy participants (M3). A model assuming equal rates of failure between healthy and psychiatric participants was 2.88 times more likely than the model assuming greater rates for psychiatric patients. In turn, a model assuming lower rates of failure for psychiatric patients was 1.27 times more likely than the model assuming equal rates. Finally, a model assuming lower rates of failure for psychiatric patients was 3.65 times more likely than the model assuming higher rates for psychiatric patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a difference in model fit as meaningful. Though the small sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data is the one where psychiatric patients are more likely to fail infrequency item attention checks. These data suggest that it is unlikely that participants flagged for C/IE responding in the main analyses were disproportionately individuals with high psychiatric symptom severity. Accordingly, we conclude that our results are unlikely to be overly influenced by an overcontrol bias.

@@

In this manuscript, we highlighted a particular set of circumstances, common in computational psychiatry research done on large online samples, in which spurious correlations may arise between task behavior and self-reported symptomology. When the ground-truth prevalence of a symptom is low in the general population, participants who respond carelessly on measures assessing this symptom may erroneously appear as symptomatic. A less careful pattern of responding on tasks used to measure cognitive constructs can then masquerade as a correlation between alteration in these constructs and symptom dimensions. We found empirical evidence for this pernicious pattern in participants recruited from two popular online labour platforms. False-positive rates for these spurious correlations \emph{increased} with sample size, because the correlations are due to measurement bias, not measurement noise. Importantly, we found that screening on task behavior alone was insufficient to identify participants engaging in C/IE responding and avoid the false-positive correlations. Unfortunately, a literature review identified this type of screening as the most common practice in online computational psychiatry studies. We recommend instead to screen and exclude participants based on responding on surveys, a practice that abolished many spurious behavior-symptom correlations in our data.

@@

Viewed from the perspective of effort expenditure, our results suggest that participants appraised the cost/benefit trade-off differently for behavioral tasks and self-report surveys. Specifically, we found that only 7\% of participants were at chance-level performance in our task, compared to more than 22\% of participants who failed one or more attention-check items in the self-report surveys. Moreover, different measures of C/IE responding were weakly or not at all correlated between task behavior and self-report responses. This suggests the motivation for effortful responding was greater in the behavioral task, though precisely why is unclear. One possibility is that we gave participants a monetary incentive for attentive responding only during the task (a common practice, according to our literature review). A second possibility is that participants expected fewer consequences for C/IE responding during the self-report surveys, a reasonable assumption in light of how infrequently previous experiments have screened self-report data. Alternatively, participants may have found the gamified behavioral task more engaging or the self-report inventory more tedious. Regardless of the reason, this discrepancy reinforces our observations concerning the inadequacy of behavioral-task screening as a stand-alone method for identifying C/IE responding. Since, in general, participants may appraise costs and benefits of effortful responding differently for behavioral tasks and self-report surveys, screening for C/IE responding on one data modality may in general be unsuitable for identifying it in the other. We therefore recommend screening on each component of an experiment.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g. low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence for the proposition that patients were more likely to fail attention checks than healthy controls (if anything, non-psychiatric participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias.

Given that the results of our patient study are preliminary and warrant further investigation, researchers might still be wary of possible overcontrol bias. However, when using self-report questionnaires for screening, for overcontrol to seriously impact results it would have to be the case that symptomatic participants frequently endorse improbable or impossible responses to infrequency-item checks (e.g., responding `Agree' to ``I competed in the 1917 Olympic Games"). In this case, and even if such participants truly are experiencing severe symptoms of motivation or attention, there is likely to be limited utility in measuring these symptoms using a self-report measure that they are unable to complete veridically. A similar rationale underlies the widespread use of semi-structured interviews and other clinician-report measures rather than self-report measures for in-clinic psychiatric research. We would therefore argue that, if the psychiatric phenomenon being studied is such that this issue warrants concern, the research question may be better suited to an in-person study design involving participants in the clinic who meet full diagnostic criteria than a correlational design involving an online convenience sample.

>>>>

% load packages
\usepackage[utf8]{inputenc}
\usepackage[a4paper, margin=0.8in]{geometry}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{xr-hyper} % for cross-referencing between documents
\usepackage{hyperref}
\usepackage{float}
\usepackage{amsmath}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[flushmargin]{footmisc}
\usepackage[flushleft]{threeparttable}
\usepackage{array}
\usepackage[font=small, labelfont=it]{caption}
\usepackage{booktabs}
\usepackage{xcolor,soul}

% define highlight style
% \colorlet{usercolorname}{yellow!00}    % use for public version
\colorlet{usercolorname}{yellow!30}    % use for reviewer version
\sethlcolor{usercolorname}

@@

% define highlight style
% \colorlet{usercolorname}{yellow!00}    % use for public version
\colorlet{usercolorname}{yellow!30}    % use for reviewer version
\sethlcolor{usercolorname}

% specify images path relative to the main .tex file 
\graphicspath{ {./figures/} }

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1]{Samuel Zorowitz}
\author[2]{Johanne Solis}
\author[1,3]{Yael Niv}
\author[4]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, NJ, USA}
\affil[2]{Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry, Rutgers University, NJ, USA}
\affil[3]{Department of Psychology, Princeton University, NJ, USA}
\affil[4]{School of Psychological Sciences, Monash University, Victoria, Australia}

@@

In recent years, online labour markets (e.g. Amazon Mechanical Turk, Prolific, CloudResearch) have become increasingly popular as a source of research participants in the behavioral sciences \cite{stewart2017crowdsourcing}, in no small part due to the ease with which these services allow for recruitment of large, diverse samples. The advantages of online data collection have also begun to be recognized in psychiatric research \cite{chandler2016conducting}, where this method offers several distinct advantages over traditional approaches to participant recruitment. The ability to assess psychiatric symptom severity in large general-population samples makes possible large-scale transdiagnostic analysis \cite{gillan2016taking, rutledge2019machine}, and facilitates recruitment from difficult-to-reach participant populations \cite{strickland2019use}. Online labour markets also facilitate re-recruitment, making them an attractive option for validating the psychometric properties of assessment tools \cite{enkavi2019large} or studying clinical processes longitudinally \cite{kothe2019retention}.

@@

Here we wish to draw special attention to an underappreciated feature of psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey \hl{attentively} \cite{chandler2020participant, ophir2020turker, king2018random}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. \hl{In two new datasets} from two separate online labour marketplaces, we found that screening based on task behavior fails to completely identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept datasets that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms that are abolished when data are screened more thoroughly.

@@

We hypothesise that spurious behavior-symptom correlations may emerge due to a mean-shift in the average level of symptom endorsement in participants engaging in C/IE responding relative to attentive participants. In turn, a mean-shift is expected to occur when the overall rate of symptom endorsement is low; \hl{that is, comparably higher scores are more likely for C/IE participants responding at random on a right skewed questionnaire.} In line with our predictions, the average level of symptom endorsement was noticeably exaggerated in C/IE-responding participants for the symptom measures where symptom scores were most positively-skewed (7-up, 7-down, GAD-7; see Figure \ref{fig:distributions}). In contrast, where there was higher rates of symptom endorsement overall, the distributions of symptom scores between the two groups of participants were less noticeably distinct. Permutation testing confirmed that observed mean-shifts in symptom scores for C/IE participants were statistically significant for the majority of symptom measures (Table \ref{tab:validation}).

@@

Second, we quantified the degree to which participants responded to self-report symptom surveys in a stereotyped fashion; that is, we determined if participants exhibited patterns in their responses that were independent of the contents of the survey items. We fit a random-intercept item factor-analysis model \cite{maydeu2006random} to self-report data (see Methods), and for each participant we estimated an intercept parameter that quantified their bias towards using responses on the left or right side of the response scale, regardless of what that response signifies for a particular self-report measure (e.g., low on one symptom scale versus high on another). We observed a credible difference between the average value of this intercept for the two groups ($\Delta \text{intercept} = 0.67, 95\% \ \text{HDI} = [-0.78, -0.55]$), such that C/IE participants were biased towards using the right-half of survey response options. This translates to a tendency to endorse \emph{more severe} symptoms on the 7-up/7-down and GAD-7 scales (where the rightmost options indicate greater frequency of symptoms) but \emph{less extreme} symptoms or \hl{personality traits on the SHAPS and BIS (where the rightmost options indicate lower frequency of symptoms or personality traits)} despite these inventories measuring strongly correlated constructs (i.e. depression and anhedonia, anxiety and behavioral inhibition).

@@

\begin{figure}[!t]
\includegraphics[width=17cm]{../figures/main_03a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. Only statistically significant correlations are shown ($p<0.05$ not corrected for multiple comparisons; \hl{all signed correlations are shown in the Supplementary Figure} \ref{fig:correlations_s1}). Black Xs indicate significant correlations ablated under screening. No Screening = no exclusions; Accuracy Only = exclusions based on chance-level performance in the reversal-learning task; Infrequency Only = exclusions based on invalid or improbable responses to infrequency items; Both Types = exclusions based on the previous two measures. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
\label{fig:correlations}
\end{figure}

@@

Next, we investigated how spurious correlations depended on sample size. To do so, we performed a bootstrapping analysis where we held fixed the proportion of participants engaging in C/IE responding (i.e. 5\%, 10\%, 15\%, 20\%) and increased the total number of participants. Across all analyses, we measured the correlation and between the 7-down depression scale and learning-rate asymmetry ($\kappa$), which we previously identified as likely exhibiting a spurious association. (\hl{The following results are not specific to learning-rate asymmetry and generalize to other pairs of variables; Figure} \ref{fig:bootstrap_s1}).

The outputs of the bootstrapping analysis are presented in Figure \ref{fig:bootstrap}. We found that, although estimated correlation magnitudes were independent of sample size (x-axis, left panel), the absolute magnitude of the behavior-symptom correlation increased with the proportion of C/IE participants (different coloured circles, left panel). Crucially, we found false-positive rates for spurious correlations \emph{increased} with increases in sample size in our data for all but the smallest rates of C/IE responding (right panel). This runs counter to a common assumption that larger sample sizes are protective against spurious correlations because they serve to mitigate measurement error. Although this assumption is correct for unsystematic measurement error, it no longer holds in the regime of systematic measurement error (where larger sample sizes reduce the variance of estimates, but do not alter their bias). Instead, our results suggest that, \hl{except for low rates of C/IE responding}, the false-positive rate for behavioral-symptom correlations will become increasingly inflated as the sample size increases.

@@

\subsection{\texorpdfstring{\hl{Pattern of results generalizes to alternative tasks, self-report measures, and quality assurance protocols}}{}}

One possible concern with the results presented so far is that they are specific to one instantiation of our experimental design. With more stringent quality assurance protocols during participant recruitment, or perhaps a different task or set of self-report measures, one might wonder if spurious correlations would remain such a threat.

To evaluate the generalizability of our findings, we therefore conducted a conceptual replication experiment in which an independent sample of N=393 participants (N=193 from MTurk using CloudResearch, N=200 from Prolific) completed a more difficult cognitive task, the well-known ``two-step task'' \cite{daw2011model}, and an alternate set of self-report measures (see Appendix B for details). Importantly, participants were recruited \textit{after} CloudResearch and Prolific implemented new protocols to improve data quality on their respective platforms. As a final control measure, participants completed self-report symptom measures as before, but also personality measures with no hypothesized relationship to model-based planning behavior on the two-step task.

For the sake of brevity, we report here only the main pattern of findings (all results are reported in Appendix B). In the replication sample, 55 out of 393 participants (14\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report measures. This is roughly two-thirds of the number of participants who were flagged for C/IE responding in the original study, suggesting that the newer quality assurance protocols used by the online platforms are at least partially effective.

In the self-report symptom measures, we replicated the finding that total scores were noticeably exaggerated in participants suspected of C/IE responding, but only for symptom measures where overall rates of symptom endorsement were the lowest (i.e., measures that have normative endorsements skewed towards 0). Similarly, we again found that task-based screening and self-report screening measures showed low correspondence; that is, excluding participants on the basis of poor behavioral performance would not have identified and removed participants who engaged in C/IE responding on self-report surveys.

Finally, when we did not apply any exclusions, we observed spurious correlations between performance on the two-step task and total scores for both symptom and personality self-report measures with a mean-shift in scores between attentive participants and participants suspected of C/IE responding. In contrast with our original findings, however, we found that excluding participants based on self-report \textit{or} task screening measures was sufficient to ablate these spurious correlations.

In sum, we replicated most of the main findings from the original study in an independent sample of participants completing a different task and other self-report measures. Although we found that screening on task behavior was sufficient to protect against spurious correlations in the replication sample, it is difficult to generalize and predict when or why this might be the case for other datasets. As such, we still believe that screening for C/IE responding in both task and self-report measures is the best approach to protect oneself against the possibility of spurious correlations.

\subsection{\texorpdfstring{\hl{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}}{}}

One major concern with performing rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias \cite{elwert2014endogenous}. That is, to this point we have treated the tendency towards C/IE responding as independent from psychopathology. However, to the extent that C/IE responding reflects lack of motivation \cite{barch2015mechanisms}, avoidance of effort \cite{cohen2001impairments, culbreth2016negative}, or more frequent lapses of attention \cite{kane2016individual, robison2017neurotic}, one might hypothesise a true underlying association between psychopathology and careless responding in online studies. It is thus plausible that rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants.

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Appendix C for details). Specifically, N=45 psychiatric patients (confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview) and N=20 healthy controls, all recruited through the Rutgers-Princeton Center for Computational Cognitive NeuroPsychiatry (i.e., personal recruitment, not on online platforms), completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) psychiatric participants were flagged for C/IE responding.

Using these data, we computed pairwise Bayes factors comparing three candidate models: attention check failure rates are equal between healthy and psychiatric participants (M1); failure rates are greater in psychiatric patients (M2); and failure rates are greater in healthy participants (M3). The model assuming equal rates of failure between healthy and psychiatric participants was 2.88 times more likely than the model assuming greater rates for psychiatric patients. In turn, the model assuming lower rates of failure for psychiatric patients was 1.27 times more likely than the model assuming equal rates. Finally, the model assuming lower rates of failure for psychiatric patients was 3.65 times more likely than the model assuming higher rates for psychiatric patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a difference in model fit as meaningful. Although the size of the sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where psychiatric patients are more likely to fail infrequency item attention checks. These data suggest, therefore, that it is unlikely that individuals with high psychiatric symptom severity were disproportionately flagged for C/IE responding in the main analyses. Accordingly, we conclude that the screening measures we are suggesting are not likely to result in overcontrol bias and false-negative correlations between tasks and symptom measures.

@@

In this manuscript, we highlighted a particular set of circumstances, common in computational psychiatry research done on large online samples, in which spurious correlations may arise between task behavior and self-reported symptomology. When the ground-truth prevalence of a symptom is low in the general population, participants who respond carelessly on measures assessing this symptom may erroneously appear as symptomatic. A less careful pattern of responding on tasks used to measure cognitive constructs can then masquerade as a correlation between alteration in these constructs and symptom dimensions. We found repeated evidence for this pernicious pattern in participants recruited from two popular online labour platforms. False-positive rates for these spurious correlations \emph{increased} with sample size, because the correlations are due to measurement bias, not measurement noise. Importantly, we found that screening on task behavior alone was insufficient to identify participants engaging in C/IE responding and avoid the false-positive correlations. Unfortunately, a literature review identified this type of screening as the most common practice in online computational psychiatry studies. We recommend instead to screen and exclude participants based on responding on surveys, a practice that abolished many spurious behavior-symptom correlations in our data.

@@

Viewed from the perspective of effort expenditure, our results suggest that participants appraised the cost/benefit trade-off differently for behavioral tasks and self-report surveys. Specifically, we found that only 7\% of participants were at chance-level performance in our task, compared to more than 22\% of participants who failed one or more attention-check items in the self-report surveys \hl{(a finding that replicated in a second study involving a different task)}. Moreover, different measures of C/IE responding were weakly or not at all correlated between task behavior and self-report responses. This suggests the motivation for effortful responding was greater in the behavioral task, though precisely why is unclear. One possibility is that we gave participants a monetary incentive for attentive responding only during the task (a common practice, according to our literature review). A second possibility is that participants expected fewer consequences for C/IE responding during the self-report surveys, a reasonable assumption in light of how infrequently previous experiments have screened self-report data. Alternatively, participants may have found the gamified behavioral task more engaging or the self-report inventory more tedious. Regardless of the reason, this discrepancy reinforces our observations concerning the inadequacy of behavioral-task screening as a stand-alone method for identifying C/IE responding. Since, in general, participants may appraise costs and benefits of effortful responding differently for behavioral tasks and self-report surveys, screening for C/IE responding on one data modality may in general be unsuitable for identifying it in the other. We therefore recommend screening on each component of an experiment.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g. low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. \hl{To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence for the proposition that patients were more likely to fail attention checks than healthy controls (if anything, non-psychiatric participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias.}

\hl{Given that the results of our patient study are preliminary and warrant further investigation, researchers might still be wary of possible overcontrol bias.} However, when using self-report questionnaires for screening, for overcontrol to seriously impact results it would have to be the case that symptomatic participants frequently endorse improbable or impossible responses to infrequency-item checks (e.g., responding `Agree' to ``I competed in the 1917 Olympic Games"). In this case, and even if such participants truly are experiencing severe symptoms of motivation or attention, there is likely to be limited utility in measuring these symptoms using a self-report measure that they are unable to complete veridically. A similar rationale underlies the widespread use of semi-structured interviews and other clinician-report measures rather than self-report measures for in-clinic psychiatric research. We would therefore argue that, if the psychiatric phenomenon being studied is such that this issue warrants concern, the research question may be better suited to an in-person study design involving participants in the clinic who meet full diagnostic criteria than a correlational design involving an online convenience sample.

Notwithstanding the above, one response to this legitimate concern is to take a graded approach to screening and excluding participants \cite{Kim2018-ev}. That is, participants could be screened with respect to a multitude of measures and only the consistently flagged participants be removed, thereby reducing the risk of inducing bias. Another possibility is to use sensitivity analysis as an alternative to exclusion, testing whether full-sample observed correlations are robust to the exclusion of participants flagged by measures of C/IE responding. We note that the strict screening approach used in the present study did not preclude us from identifying symptomatic participants or behavior-symptom correlations. Indeed, we found in our sample roughly 10\% of participants endorsing symptoms consistent with clinical levels of depression, and approximately 20\% consistent with clinical levels of acute anxiety. These estimates are within the realm of epidemiological norms \cite{kessler2012twelve, lowe2008validation, yarrington2021impact}. \hl{(We should note, however, that some studies have found elevated rates of psychiatric symptomology in online participants even after controlling for C/IE responding} \cite{ophir2020turker}.) We also observed some positive correlations between anxiety and choice behavior that were consistent with effects found in previous literature \cite{huang2017computational, harle2017anhedonia, garrett2018updating}. For example, we found higher lose-shift rates and higher learning rates following negative prediction errors correlated with self-reported anxiety. This suggests that the screening methods we employed were not so aggressive as to attenuate behavior-symptom correlations that would be expected from the literature.
